Learning a functional distance between organizations with unsupervised embeddings of scientific mobility

### 2020-07-14 paper/current/main.tex
minor edits, throughout methods and supplementals

%
% Figure - SemAxis results
%
\begin{figure}[hp!]
	\centering
	\includegraphics[width=\textwidth]{\figdir/SemAxis/semaxis_fig.pdf}
	\caption{
		\textbf{Embedding captures latent geography and prestige hierarchy.}
		\textbf{a.} Illustrates SemAxis projection along two axes,California and Massachusetts and for the top five-to geographically-matched bottom five universities ranked by the 2018 Times Higher Education ranking of U.S. Universities.
		Points correspond to organizations, and a sample of states are highlighted.
		Full organization names listed in Table~\ref{table:supp:orglabels}.
		\textbf{b.} Comparison between the ranking of organizations by their rank in the Times ranking and the SemAxis rank derived from the similarity to the \textit{Elite} end of the axis prestige axis \textit{a}.
		White points are those top and bottom five universities used to define the poles of the axis.
	  	Even when considering only a total of ten organization vectors, the estimate of the spearman rank correlation between the SemAxis and Times ranking is 0.73 (n = 145), and increases if controlling for the number of top-and-bottom ranked universities (Fig.~\ref{fig:supp:semaxis_compare}).
	}
	\label{fig:semaxis}
\end{figure}

@@

Scientific mobility is complex and can be defined in many possible ways~\autocite{robinson2019mobility} and at multiple levels of analysis.
Here, authors are classified as mobile when they have at least two distinct organization ids in their trajectory, meaning that they have published using two or more distinct affiliations between 2008 and 2019.
Under this definition, mobile authors constitute 3,007,192 or 23.2 percent of all authors, and 17,700,095 author-affiliation combinations.
Mobile authors were associated with, on average, 2.5 distinct organizational affiliations.
Rates of mobility tended to differ across country.
For example, France, Qatar, the USA, Iraq, and Luxembourg had the most mobile authors (Fig.~\ref{fig:supp:descriptives}c).
However, due to their size, the USA, accounted for nearly 40 percent of all mobile authors worldwide (Fig.~\ref{fig:supp:descriptives}a), with 10 countries accounting for 80 percent of all mobility (Fig.~\ref{fig:supp:descriptives}b).
In most cases, countries with a high degree inter-organization mobility also had a high degree of international mobility;
however, the USA and France had a high degree of inter-organization mobility with the corresponding international mobility (Fig.~\ref{fig:supp:descriptives}c).
While the number of publications has increased year-to-year, the mobility and disciplinary makeup of the dataset has not significantly changed across the period of study (Fig.~\ref{fig:supp:pubs_over_time}).

\subsection*{Embedding}
%% Mobility sentences
We embed co-affiliation trajectories by treating them analogously to sentences, and organizations analogously to words.
The trajectories of mobile authors were converted to ''mobility sentences`` by concatenating together their ordered list of unique organization identifiers, as demonstrated in Fig.~\ref{fig:methods:mobility_sentence}.a.
In more complex cases, such as listing multiple affiliations on the same paper or publishing with different affiliations on multiple affiliations in the same year, the order is randomized within that year, as shown in  Fig.~\ref{fig:methods:mobility_sentence}.b.

%% word2vec
These mobility sentences are used as input to the standard skip-gram negative sampling model, for word embedding, commonly known as \textit{word2vec}~\autocite{mikolov2013word2vec} .
In natural-language processing, word embeddings are used to construct dense vector-space representations of words and phrases, in which distance relationships between words correspond to a notion of semantic distance.
By embedding co-affiliation trajectories, we aim to learn a dense vector for every organization where distance between vectors relates to the tendency for two organizations to co-occur in a token window around the target term in the same career trajectory.
Vectors are learned by training a neural network to optimize the prediction of the so-called context term given a target, in our case using one organization to predict the others appearing in the same co-affiliation trajectory.
The hidden layer of the trained neural network is used as the

@@

In embedding space, we can consider two proximity measures: cosine similarity and dot product. Cosine similarity only consider about angle difference of two embedding vectors, while dot product cares about angle and magnitude at the same time. If vectors are normalized, they are the same. In general, which measure is good depends on the task. If the task is desirable to ignore the magnitude, then cosine similarity is better, otherwise, the dot product is better.

@@

We also calculate two forms of distance between organizations.
The geographic distance, $g_{ij}$, is the pairwise geographic distance between organizations.
Geographic distance is calculated as the great circle distance, in kilometers, between pairs of organizations.
When less than one kilometer, a distance of one is imputed.
The embedding distance with the cosine similarity, $d_{ij}$, is calculated as the pairwise cosine distance between vectors from the learned organization embedding, where $v_{i}$ and $v_{j}$ represent the vectors and $d_{ij} = 1 - \frac{v_{i} \cdot v_{j}}{\lVert v_{i} \rVert \lVert v_{j} \rVert}$.
Additionally, the embedding distance with the dot product, we use  $d_{ij} =  -v_{i} \cdot v_{j}$.

@@

We consider separate variants of $f(r_{ij})$ for geographic distance, $g_{ij}$, and the embedding distance, $d_{ij}$, report the best-fit model of each distance.
For geographic distance, we use the power-law function of the gravity law, with $f(g_{ij})= g^{-\alpha}_{ij}$, where $\alpha$ is estimated from the data.
For embedding distance , we use the exponential function, with $f(d_{ij}) = e^{-\beta d_{ij}}$, where $\beta$ is estimated from the data.
To estimate parameters $C$, $\alpha$ and $\beta$ are estimated by log-transforming each variant of \ref{eq:gravity_basic}, obtaining Eq.~\ref{eq:linear_flux_geo} and Eq.~\ref{eq:linear_flux_emb} in the form of linear equations, and fitting parameters using linear regression.

\begin{equation}
	\label{eq:linear_flux_geo}
	\ln(\frac{T_{ij}}{m_im_j}) = \ln(C)  - \alpha \ln(g_{ij})
\end{equation}

\begin{equation}
	\label{eq:linear_flux_emb}
	\ln(\frac{T_{ij}}{m_im_j}) = \ln(C) - \beta d_{ij}
\end{equation}

@@

%
% SUBSECTION: SEMAXIS
%
\subsection*{SemAxis}
By constructing semantic embeddings from mobility trajectories, we allow for unique methodology for understanding the embedding space, such as SemAxis.
SemAxis~\autocite{an2018semaxis} is a technique that leverages the latent semantic characteristics of word embeddings to represent the position of terms along a conceptual axis, reflecting the relationship of these term to the concept.
The precise axis is an antonym pair defined \textit{ad-hoc} for the specific use case.
In the case of language, the ends of the axis could be ``good'' and ``bad'', ``surprising'' and ``unsurprising'', or ``masculine'' and ``feminine''. It is shown that projection of words to given specific axis fit well with widely shared cultural connotation from surveyed
responses and labeled historical data \autocite{kozlowski2018geometry}.

In this analysis, we embed organization identifiers and trajectories, rather than words and sentences, however mobility can be thought of semantically, just like language. 
Under this view, organizations can be thought of as similar when they have similar \textit{contexts}--that is, authors affiliated with that organization also tend to co-affiliate the same set of other organizations.
By treating organizations analogously to words and trajectories analogously to sentences, we can use SemAxis to leverage the semantic properties of the resulting embedding vectors to operationalize abstract relationships between organizations.

Here, we apply SemAxis to U.S. universities by defining two axes, one of geography and one of academic prestige.
The poles of the geographic axis are defined as the mean vector of all vectors corresponding to organizations in California, and then the mean of all vectors of organizations in Massachusetts.
For the prestige axis, we define a subset of top-ranked universities according to either the Times World University Rankings or based on the mean normalized research impact sourced from the Leiden Rankings.
The other end of the prestige axis is the geographically-matched (according to census region) set of universities ranked at the bottom of these rankings.
For example, if 20 universities top-ranked universities are selected and six of them are in the Northeastern United States, then the bottom six universities according to the same ranking will be included in the low-ranked subset.
From the prestige axis we also derive what we call \textit{SemRank}, a university ranking based on the nearness of universities to the \textit{Elite} end of the axis.
We compare the SemAxis ranking to the traditional university rankings using Spearman rank correlations.

@@

\begin{equation}
	\label{eq:KL}
	D_{KL} = \sum p(x)\log(\frac{p(x)}{q(x)})
\end{equation}

@@

%
% Figure - SemAxis results
%
\begin{figure}[hp!]
	\centering
	\includegraphics[width=\textwidth]{\figdir/SemAxis/semaxis_compare_fig.pdf}
	\caption{
		\textbf{SemRank hierarchy is robust.}
		\textbf{a.} Spearman correlation (n = 143) between Times prestige rank and SemAxis rank, by the number of high and bottom-ranked (geographically matched) universities aggregated to form the poles.
		Black points show spearman correlation using all organizations; white points show correlation using only universalizes not aggregated in the poles. 
		More universities improves performance, but quickly saturates after around five universities.
		\textbf{b - f.} Comparison between the Times and SemAxis ranks of universities, by number of universities used to define the poles (n).
		White points are those top and bottom 20 universities aggregated to define the ends of the axis.
		The grey box corresponds to the top 20 and bottom 20 ranks.
	  	PCC details the estimate from Spearman correlation between the two rankings.
	}
	\label{fig:supp:semaxis_compare}
\end{figure}

>>>>

%
% Figure - SemAxis results
%
\begin{figure}[hp!]
	\centering
	\includegraphics[width=\textwidth]{\figdir/SemAxis/semaxis_fig.pdf}
	\caption{
		\textbf{Embedding captures latent geography and prestige hierarchy.}
		\textbf{a.} Illustrates SemAxis projection along two axes,California and Massachusetts and for the top five-to geographically-matched bottom five universities ranked by the 2018 Times Higher Education ranking of U.S. Universities.
		Points correspond to organizations, and a sample of states are highlighted.
		Full organization names listed in Table~\ref{table:supp:orglabels}.
		\textbf{b.} Comparison between the ranking of organizations by their rank in the Times ranking and the SemAxis rank derived from the similarity to the \textit{Elite} end of the axis prestige axis \textit{a}.
		White points are those top and bottom five universities used to define the poles of the axis.
	  	Even when considering only a total of ten organization vectors, the estimate of the Spearman's rank correlation between the SemAxis and Times ranking is 0.73 (n = 145), and increases if controlling for the number of top-and-bottom ranked universities (Fig.~\ref{fig:supp:semaxis_compare}).
	}
	\label{fig:semaxis}
\end{figure}

@@

Scientific mobility is complex and can be defined in many possible ways~\autocite{robinson2019mobility} and at multiple levels of analysis.
Here, authors are classified as mobile when they have at least two distinct organization ids in their trajectory, meaning that they have published using two or more distinct affiliations between 2008 and 2019.
Under this definition, mobile authors constitute 3,007,192 or 23.2 percent of all authors, and 17,700,095 author-affiliation combinations.
Mobile authors were associated with, on average, 2.5 distinct organizational affiliations.
Rates of mobility tended to differ across country.
For example, France, Qatar, the USA, Iraq, and Luxembourg had the most mobile authors (Fig.~\ref{fig:supp:descriptives}c).
However, due to their size, the USA, accounted for nearly 40 percent of all mobile authors worldwide (Fig.~\ref{fig:supp:descriptives}a), with 10 countries accounting for 80 percent of all mobility (Fig.~\ref{fig:supp:descriptives}b).
The countries with the highest proportion of mobile scientists are France, Qatar, the United States, and Iraq, whereas those with the lowest are Jamaica, Serbia, Bosnia \& Herzegovina, and North Macedonia\ref{fig:supp:descriptives}.c).
In most cases, countries with a high degree of inter-organization mobility also have a high degree of international mobility, indicating a high proportion of their total mobility being international (Fig.~\ref{fig:supp:descriptives}.d);
However, some countries such as France and the United States seem to have more domestic mobility than international mobility. 
While the number of publications has increased year-to-year, the mobility and disciplinary makeup of the dataset has not significantly changed across the period of study (Fig.~\ref{fig:supp:pubs_over_time}).

\subsection*{Embedding}
%% Mobility sentences
We embed co-affiliation trajectories by treating them analogously to sentences, and organizations analogously to words.
The trajectories of mobile authors were converted to ``affiliation trajectories'' by concatenating together their ordered list of unique organization identifiers, as demonstrated in Fig.~\ref{fig:methods:mobility_sentence}.a.
In more complex cases, such as listing multiple affiliations on the same paper or publishing with different affiliations on multiple affiliations in the same year, the order is randomized within that year, as shown in  Fig.~\ref{fig:methods:mobility_sentence}.b.

%% word2vec
These affiliation trajectories are used as input to the standard skip-gram negative sampling model, for word embedding, commonly known as \textit{word2vec}~\autocite{mikolov2013word2vec} .
In natural-language processing, word embeddings are used to construct dense vector-space representations of words and phrases, in which distance relationships between words correspond to a notion of semantic distance.
By embedding affiliation trajectories, we aim to learn a dense vector for every organization where distance between vectors relates to the tendency for two organizations to co-occur in a token window around the target organization token in the same career trajectory.
Vectors are learned by training a neural network to optimize the prediction of the so-called context term given a target, in our case using one organization to predict the others appearing in the same co-affiliation trajectory.
The hidden layer of the trained neural network is used as the final representation.

@@

@@

We also calculate two forms of distance between organizations.
The geographic distance, $g_{ij}$, is the pairwise geographic distance between organizations.
Geographic distance is calculated as the great circle distance, in kilometers, between pairs of organizations.
When less than one kilometer, a distance of one is imputed.
The embedding distance with the cosine similarity, $d_{ij}$, is calculated as the pairwise cosine distance between vectors from the learned organization embedding, where $v_{i}$ and $v_{j}$ represent the vectors and $d_{ij} = 1 - \frac{v_{i} \cdot v_{j}}{\lVert v_{i} \rVert \lVert v_{j} \rVert}$.

@@

We consider separate variants of $f(r_{ij})$ for geographic distance, $g_{ij}$, and the embedding distance, $d_{ij}$, report the best-fit model of each distance.
For geographic distance, we use the power-law function of the gravity law, with $f(g_{ij})= g^{-\alpha}_{ij}$, where $\alpha$ is estimated from the data.
For embedding distance, we use the exponential function, with $f(d_{ij}) = e^{-\beta d_{ij}}$, where $\beta$ is estimated from the data.
To estimate parameters $C$, $\alpha$ and $\beta$ are estimated by log-transforming each variant of \ref{eq:gravity_basic}, obtaining Eq.~\ref{eq:linear_flux_geo} and Eq.~\ref{eq:linear_flux_emb} in the form of linear equations, and fitting parameters using linear regression.

\begin{equation}
	\label{eq:linear_flux_geo}
	\ln\frac{T_{ij}}{m_im_j} = \ln(C)  - \alpha \ln(g_{ij})
\end{equation}

\begin{equation}
	\label{eq:linear_flux_emb}
	\ln\frac{T_{ij}}{m_im_j} = \ln(C) - \beta d_{ij}
\end{equation}

@@

%
% SUBSECTION: SEMAXIS
%
\subsection*{SemAxis}
SemAxis~\autocite{an2018semaxis} is a technique that leverages the latent semantic characteristics of word embeddings to represent the position of terms along a conceptual axis, reflecting the relationship of these term to the concept.
Each axis is an antonym pair defined for each specific use case.
In the case of language, the poles of the axis could be ``good'' and ``bad'', ``surprising'' and ``unsurprising'', or ``masculine'' and ``feminine''. 
By treating organizations analogously to words and trajectories analogously to sentences, we can use SemAxis to leverage the semantic properties of the resulting embedding vectors to operationalize abstract relationships between organizations.

Given vectors representing the positive and negative poles of an axis, $v^{+}$ and $v^{-}$, the axis is defined as.$v_{\text{axis}} = v^{+} - v^{-}$. For each organization in the embedding space, $w$, a score is calculated relating the term to to the positive, $v^{+}$ or negative, $v^{-}$ pole, calculated as the cosine similarity of the organization's vector with the axis,

$$
score(w)v_{\text{axis}} = \cos(u_{w}, v_{\text{axis}})
$$

We apply SemAxis to U.S. universities by defining two axes, one of geography and one of academic prestige.
The poles of the geographic axis are defined as the mean vector of all vectors corresponding to organizations in California, and then the mean of all vectors of organizations in Massachusetts.
For the prestige axis, we define a subset of top-ranked universities according to either the Times World University Rankings or based on the mean normalized research impact sourced from the Leiden Rankings.
The other end of the prestige axis is the geographically-matched (according to census region) set of universities ranked at the bottom of these rankings.
For example, if 20 universities top-ranked universities are selected and six of them are in the Northeastern United States, then the bottom six universities according to the same ranking will be included in the low-ranked subset.
From the prestige axis we derive a ranking of universities that we then compare to other formal university rankings using Spearman rank correlation.

@@

\begin{equation}
	\label{eq:KL}
	D_{KL} = \sum p(x)\log\frac{p(x)}{q(x)}
\end{equation}

@@

%
% Figure - SemAxis results
%
\begin{figure}[hp!]
	\centering
	\includegraphics[width=\textwidth]{\figdir/SemAxis/semaxis_compare_fig.pdf}
	\caption{
		\textbf{SemRank hierarchy is robust.}
		\textbf{a.} Spearman correlation (n = 143) between Times prestige rank and SemAxis rank, by the number of high and bottom-ranked (geographically matched) universities aggregated to form the poles.
		Black points show spearman correlation using all organizations; white points show correlation using only universalizes not aggregated in the poles. 
		More universities improves performance, but quickly saturates after around five universities.
		\textbf{b - f.} Comparison between the Times and SemAxis ranks of universities, by number of universities used to define the poles (n).
		White points are those top and bottom 20 universities aggregated to define the ends of the axis.
		The grey box corresponds to the top 20 and bottom 20 ranks.
	  	Spearman's rho details the estimate from Spearman correlation between the two rankings using all universities, including those used to define the ends of each axis. 
	}
	\label{fig:supp:semaxis_compare}
\end{figure}
